### 2024-09-23 Analysis of bias in arrests/paper/paper.qmd
Initial Commit

---
title: "No Evidence of Bias in Arrests Made by the Toronto Police Service"
author: "Aviral Bhardwaj"
abstract: The ethnic identity of citizens arrested by police has consistently sparked discussion, highlighting skewness in certain ethnicities. This paper analyzes Toronto Police Service arrest data from 2020 to 2021, considering factors such as age and search on arrest. While population distribution varies significantly, findings suggest no inherent bias in arrests. However, strict enforcement is consistent across all ethnicities and races, even when considering evidence of higher crime rates in specific groups. Further analysis can provide insights for policymakers, law enforcement, and community advocates, emphasizing the need for transparency and accountability in policing.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Aviral-03/Ethnic-Identity-Arrests-Toronto"
bibliography: references.bib
---

```{r}
#| echo: false
# All the required packages
## Installing Packages only needs to be done once

#install.packages("tidyverse")
#install.packages("tinytex")
#install.packages("opendatatoronto")
#install.packages("kableExtra")
#install.packages("patchwork")
#install.packages("janitor")
#install.packages("gridExtra")
```

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(tinytex)
library(knitr)
library(dplyr)
library(ggplot2)
library(janitor)
library(patchwork)
library(kableExtra)
library(gridExtra)
```

```{r}
#| warning: false
#| message: false
#| include: false

# Get package data

# get package
package <- show_package("police-race-and-identity-based-data-collection-arrests-strip-searches")
package

# get all resources for this package
resources <- list_package_resources("police-race-and-identity-based-data-collection-arrests-strip-searches")

# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))

# load the first datastore resource as a sample
data <- filter(datastore_resources, row_number()==1) %>% get_resource()

# Save the data in CSV
write.csv(data, "../../inputs/data/raw_data.csv")
```

## Introduction

In the vibrant metropolis of Toronto, the intersection between law enforcement practices and demographic characteristics holds profound implications for community dynamics and public policy. Toronto, named Canada's most diverse city, encounters numerous challenges arising from the coexistence of ethnically diverse populations within the same neighborhoods. According to an article published by Georgetown University, the "ongoing debate surrounding police reforms is heavily influenced by historical connections with slavery and contemporary racial disparities" (@rJackson2022). This statement underscores the contentious nature of whether the race identity affiliation of police officers affects their policing practices, despite the expectation for fair and unbiased treatment.

In a recent article published by the Toronto Police Services, it is highlighted that "Black, East/Southeast Asian, Middle Eastern, and Latino people were over-represented in reported use of force incidents compared to their presence in the enforcement action population" (@Fanfair2022). This indicates that, despite Toronto Police Services working towards a more equitable society, there are still hints of racial bias in the arrests made by the Toronto Police Services.Conversely this paper undertakes a comprehensive analysis of arrest data of various race and ethinic iden spanning the years 2020 and 2021 in the city of Toronto, seeking to unravel the intricate patterns that shape the city's legal landscape.

Arrest data serves as a vital lens through which we can gain insights into the dynamics of law enforcement activities and their correlation with ethnicity and age. In the data section we clean and process our data. Following which in our discussion and results our inquiry begins with a broad exploration of the distribution of arrests across various age groups, aiming to discern overarching trends and fluctuations. The age spectrum, a fundamental dimension of this analysis, reveals intriguing patterns that underscore the nuanced nature of arrests in Toronto. This paper aims to contribute to understanding the dynamics of arrests in Toronto, providing insights that can inform targeted interventions, policy considerations, and community engagement efforts.

## Data

This data is sourced from the Toronto Open Data Portal `opendatatoronto` [@rOpenDataToronto]. The following data is used for an in-depth analysis of arrests and strip searches which involve various ethnic and identity-based groups in the city of Toronto. The information is gathered by the Toronto Police Service (TPS) and is collected under the authority of the Police Services Act. Its purpose is to improve the relationship between the police and the community by being more transparent with the public. The data covers the period from 2020 to 2021.

The data underwent cleaning and analysis using the R programming language [@r]. Cleaning was performed with the `tidyverse` package [@rTidyverse], involving the removal of unnecessary columns for the analysis. Subsequently, analysis was conducted utilizing the `dplyr` package [@rDplyr], with the data grouped by ethnic and identity-based categories. Visualization was then executed using the `ggplot2` package [@rGgplot2], depicting the number of arrests and strip searches across various ethnic and identity-based groups.

The raw dataset comprised 32,000 arrest records spanning both 2020 and 2021, containing 26 variables of analyzable data. However, for the analysis in this paper, we narrowed our focus to specific variables, including Arrest Year, Perceived Race, Age Group, Youth at Arrest, Booked, Search Reason Cause Injury, Search Reason Assist Escape, Search Reason Possess Weapons, and Search Reason Possess Evidence. Additionally, we consolidated similar columns, namely Search Reason Cause Injury, Search Reason Assist Escape, Search Reason Possess Weapons, and Search Reason Possess Evidence, into a unified column labeled "Search Reason." utilizing the `any()` function [@rDplyr].

```{r}
#| warning: false
#| message: false
#| include: false

# Reading the data
data <- read.csv("../../inputs/data/raw_data.csv")

# Adjusting and cleaning the name of the variable
cleaned_data <- data %>% 
  clean_names() %>% 
  clean_names(case = "snake")

cleaned_data <- cleaned_data |>
  select(
    arrest_year, perceived_race, age_group_at_arrest, youth_at_arrest_under_18_years, booked, search_reason_cause_injury, search_reason_assist_escape, search_reason_possess_weapons, search_reason_possess_evidence)

# Combining the columns that are similar
cleaned_data <- cleaned_data %>%
  mutate(search_reason = case_when(
    search_reason_cause_injury == 1 ~ "1",
    search_reason_assist_escape == 1 ~ "1",
    search_reason_possess_weapons == 1 ~ "1",
    search_reason_possess_evidence == 1 ~ "1",
    TRUE ~ "0"
  ))

# Rename the columns
cleaned_data <- cleaned_data %>%
  rename(
    age_group = age_group_at_arrest,
    youth_at_arrest_under_18_years = youth_at_arrest_under_18_years
  )

# Mutating the date for age groups involving Aged 17 years and younger and Aged 17 years and under essentially they mean the same thing
cleaned_data <- cleaned_data %>%
  mutate(age_group = case_when(
    age_group == "Aged 17 years and younger" ~ "Aged 17 years and under",
    age_group == "Aged 17 years and under" ~ "Aged 17 years and under",
    TRUE ~ age_group
  ))

# Mutating for booked with NA values
cleaned_data <- cleaned_data %>%
  mutate(booked = case_when(
    is.na(booked) ~ "0",
    booked == 1 ~ "1",
    TRUE ~ "Not Booked"
  ))

cleaned_data <- cleaned_data |>
  select(
    arrest_year, perceived_race, age_group, youth_at_arrest_under_18_years, booked, search_reason)

write.csv(cleaned_data, "../data/cleaned_data.csv")

```

```{r}
#| warning: false
#| message: false
#| include: false
# Make a table of the data with the number of arrests and strip searches by ethinic and identity based groups for the year 2020, and 2021
# cleaned_data <- read.csv("../data/cleaned_data.csv")

# List of the ethnic and identity based groups that are in the data
# ethnic_backgrounds <- cleaned_data |> select(perceived_race) |> unique()

# List of the years that are in the data
# years <- cleaned_data |> select(arrest_year) |> unique()

# List of the age groups that are in the data
# age_groups <- cleaned_data |> select(age_group) |> unique()

# write.csv(cleaned_data, "../data/cleaned_data.csv")
```

### Arrests by Ethnic and Identity-Based Groups for Year 2020 and 2021

Since our analysis mainly focuses on the distribution of arrests for each ethnic identity, we extracted all ethnic identities from our data. These ethnic identities include Black, White, South Asian, East Asian, Middle Eastern, Latin American, Southeast Asian, and Indigenous. The data for the years 2020 and 2021 are summarized in Table 1 (a) (@tbl-table1) and Table 1 (b), respectively.

Table 1 (a) (

@tbl-table1) provides a summary of our data, revealing the number of individuals arrested from each major ethnic group in the years 2020 and 2021. Table 2 (b) shows, out of all the people who were arrested from each ethnic identity, how many of them were actually booked under a criminal case.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-table1
#| tbl-cap: "Total number of arrests vs Booked for the year 2020, and 2021"
## Table 1: Total Number of Arrests
cleaned_data <- read.csv("../data/cleaned_data.csv")

table1 <- cleaned_data |> 
  filter(!perceived_race == "None") |>
  group_by(arrest_year, perceived_race) |> 
  summarise(n = n(), .groups = "drop") |> 
  arrange(desc(n)) |> 
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

table2 <- cleaned_data |>
  filter(booked == 1) |>
  filter(!perceived_race == "None") |>
  group_by(arrest_year, perceived_race) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

knitr::kable(
  list(table1, table2),
  col.names = c("Ethinic Identity", "2020", "2021"),
  booktabs = TRUE,
  )

```

### Total number of arrests by age group for the year 2020 and 2021

It would also be important to take a look at the age group distribution of arrests. Table 2 (@tbl-table2) shows us this information. A data summary which indicates the count of different age groups of individuals from prominent ethnic groups in the years 2020 and 2021. All the age groups were the following: Aged 17 years and under, Aged 18 to 24 years, Aged 25 to 34 years, Aged 35 to 44 years, Aged 45 to 54 years, Aged 55 to 64 years, Aged 65 years and over.

Due to the extensive nature of the dataset, going through nine different age groups across eight ethnicities, summarizing the information for clarity is extremely difficult. Consequently, we have decided to showcase a subset of the data which focuses on the age group of 25 to 34 years so that it is easy to read.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-table2
#| tbl-cap: "Arrests Booked by ethinic background for the year 2020, and 2021"

## Table 3: Arrests by Age Group

cleaned_data <- read.csv("../data/cleaned_data.csv")
# 
# # Group 
table3 <- cleaned_data %>%
  filter(!(age_group == "None") & !(perceived_race == "None")) |>
  group_by(arrest_year, perceived_race, age_group) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(perceived_race)) %>%
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

table3_small <- cleaned_data %>%
  filter(age_group == "Aged 25 to 34 years") |>
  group_by(arrest_year, perceived_race, age_group) %>%
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

kable(table3_small,
      align = "c",
      linesep = "",
      col.names = c("Ethinic and Identity Based Groups", "Age Group", "2020", "2021"),
      booktabs = TRUE)

```

We would like to acknowledge that opendatatoronto [@rOpenDataToronto] provides the distribution of the population by ethnicity. However, due to the vast size of Toronto and its numerous wards, it has been difficult to compare all of the data and cross-verify it with our crime rate data comprehensively. Despite this, a more detailed analysis can be conducted if we compare specific ethnic groups with each other.

Now that the data has been cleaned, we can begin to analyze the data. The following section will provide a summary of the data, and the trends that emerge from the data. Figure 1 (@fig-fig1), Figure 2 (@fig-fig2), Figure 3 (@fig-fig3), and Figure 4 (@fig-fig4) was created using `ggplot` [@rGgplot2], displays the information that can be used as compare the trends as described above.

### Number of Arrests in year 2020-2021

With reference to Table 1 (@tbl-table1) and the graph (see @fig-fig1) the data shows that there is a non-uniform distribution of arrests across all ethnic groups, with the Black ethnic group and White group having the highest numbers of arrests in both 2020 and 2021. In addition to this, the data shows a decrease in the number of arrests for these two ethnic groups from 2020 to 2021. However, a  for the other ethnic groups, the number of arrests has increased over the same time period. This shows that both the trends are moving in opposite directions.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Number of arrests by ethinic background for the year 2020, and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig1

table1_long <- pivot_longer(table1, cols = c("2020", "2021"), names_to = "year", values_to = "value")

# Create a plot
ggplot(table1_long, aes(x = perceived_race, y = value, fill = year)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(title = "Bar Chart for the number of arrests by ethinic background for the year 2020, and 2021",
       x = "Ethnic and Identity Based Groups",
       y = "Number of Arrests",
       fill = "Year") +
  theme_minimal() +
  scale_fill_discrete(labels = c("2020", "2021")) +
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_brewer(palette = "Set1")
```

Referring to Table 1 (@tbl-table1) for the proportion of arrests that were booked, we plotted a stacked bar graph (see @fig-fig2), with y-axis representing the total percentage. The data shows that there is a consistent distribution of the number of individuals who were booked after arrests across various ethnic groups. Notably, there are many more individuals who were booked after being arrested as  compared to those who were arrested but not booked for a criminal case. In addition to this, we also observe that, across all ethnic groups, the percentage of individuals booked falls within the range of 50% to 55%, and this pattern remains consistent in the year 2021.

However the Black ethnicity has the highest number of individuals who were booked after being arrested in both the years 2020 and 2021.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Number of arrests booked by ethinic background for the year 2020, and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig2

table2_long <-
  cleaned_data |>
  select(
    arrest_year,
    perceived_race,
    booked
  ) |>
  group_by(arrest_year, perceived_race, booked) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(perceived_race)) %>%
  filter(!(perceived_race == "None")) %>%
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

table2_long <- pivot_longer(table2_long, cols = c("2020", "2021"), names_to = "year", values_to = "value")

ggplot(table2_long, aes(x = perceived_race, y = value, fill = factor(booked))) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ year, scales = "free_y") +
  labs(title = "Booked after Arrest in the year 2020 and 2021",
       x = "Perceived Race",
       y = "Percentage of Arrests Booked",
       fill = "Status") +
  theme_minimal() +
  theme(legend.position = "right") + 
  scale_fill_discrete(labels = c("Not Booked", "Booked")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Number of Arrests by Age Group for the year 2020-2021

As depicted in Figure 3 (see @fig-fig3), the data shows there is a non-uniform distribution of arrests across age groups, with the age groups of 25 to 34 years and 35 to 44 years showing the highest number of arrests in both 2020 and 2021. Furthermore, there is a notable decrease in the number of arrests for these two age groups from 2020 to 2021. However, a different trend emerges for other age groups, where the number of arrests has increased over the same time period.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Number of arrests by Age-Group for the year 2020, and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig3
# Bar plot

table3_long <- pivot_longer(table3, cols = c("2020", "2021"), names_to = "year", values_to = "value")

ggplot(table3_long, aes(x = perceived_race, y = value, fill = factor(age_group))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ year, scales = "free_y") +
  labs(title = "Bar Chart for the number of arrests by Age-Group for the year 2020, and 2021",
       x = "Perceived Race",
       y = "Number of People",
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Examining Figure 4 (see @fig-fig4), which shows the dot plot for the number of arrests by age group in the year 2020 and 2021 for the top two ethnic groups, "Black" and "White," it becomes evident that the age group 25 to 34 years represents the highest number of arrests for individuals of Black ethnicity. The trend for Black ethnicity indicates a decreasing curve from this age group onward. Howver, for individuals of White ethnicity, the peak in arrests is observed in the age group 35 to 44 years. From age 45 to 65, however, the difference in the number of arrests between both ethnicities starts to converge.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Scatter Plot for the number of arrests by Age-Group for the year 2020, and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig4

table3_ii <- table3 %>%
  filter(perceived_race %in% c("White", "Black")) %>%
  filter(!(age_group == "None")) %>%
  pivot_longer(cols = c("2020", "2021"), names_to = "year", values_to = "value")

# Create a line Graph
ggplot(table3_ii, aes(x = age_group, y = value, color = perceived_race)) +
  geom_point(size = 3) +
  facet_wrap(~ year, scales = "free_y") +
  labs(title = "Scatter Plot for the number of arrests by Age-Group for the year 2021",
       x = "Age Group",
       y = "Number of People",
       color = "Perceived Race") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Discussion

The analysis presented in this paper aimed to shed light on the potential bias in arrests made by the Toronto Police Service, specifically focusing on ethnic and identity-based groups. The findings from the data analysis provide valuable insights into the dynamics of law enforcement activities in the city.

### Lack of Inherent Bias in Arrests

Our analysis suggests no inherent bias in arrests made by the Toronto Police Service. The distribution of arrests across various ethnic and identity-based groups does not show a consistent pattern of discrimination. This observation is crucial, as it challenges the narrative surrounding racial bias in policing and suggests that strict enforcement is consistent across all ethnicities.

### Age Group Dynamics

The examination of age group dynamics in arrests provides additional nuances to the analysis. The age groups of 25 to 34 years and 35 to 44 years emerge as having the highest number of arrests in both 2020 and 2021. Interestingly, there is a noticeable decrease in arrests for these age groups from 2020 to 2021, suggesting potential shifts in law enforcement patterns.

The detailed exploration of age groups within specific ethnicities, such as the comparison between "Black" and "White" ethnicities, reveals intriguing patterns. The peak in arrests for individuals of Black ethnicity occurs in the age group of 25 to 34 years, while for individuals of White ethnicity, the peak is observed in the age group of 35 to 44 years. This divergence in age-specific arrest patterns highlights the importance of considering intersectionality in understanding policing dynamics.

### Implications for Policymaking and Community Engagement

Even though this paper provides several essential insights, we must acknowledge that it has its limitations. In order to understand the data better, a more detailed comparison of specific ethnic groups and further research is required. Law enforcement agencies and community advocates can gain from these insights by planning intervention targeted at specific parts of the population and community engagement efforts. This would also allow the policymakers to make any policy adjustments that they may feel is necessary to improve the situation. The emphasis on transparency and accountability in policing remains crucial, as highlighted in report by the Toronto Police Services Board [@rTPSB2019].

###Conclusion
In conclusion, tha analysis of the arrest data in the city of Toronto shows that despite there being concerns about racial bias. The data suggests that arrests may be driven by objective factors rather than racial profiling. The reforms brought about by the Toronto Police Service have strengthened transparency and fairness within the police force.While disparities may exist due to socio-economic factors, the data does not show any systemic bias.

## References

### 2024-09-23 Analysis of bias in arrests/scripts/02-data_cleaning.R

@@ -1,0 +1,67 @@
+#### Preamble ####
+# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
+# Author: Akshat Aneja
+# Date: 23 September 2024
+# Contact: akshat.aneja@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
+
+#### Workspace setup ####
+library(tidyverse)
+
+#### Clean data ####
+data <- read.csv("data/raw_data/raw_data.csv")
+
+cleaned_data <- data |>
+  select(
+    Arrest_Year, Perceived_Race, Age_group__at_arrest_, Youth_at_arrest__under_18_years, Booked, SearchReason_CauseInjury, SearchReason_AssistEscape, SearchReason_PossessWeapons, SearchReason_PossessEvidence)
+
+cleaned_data <- cleaned_data %>%
+  mutate(search_reason = case_when(
+    SearchReason_CauseInjury == 1 ~ "1",
+    SearchReason_AssistEscape == 1 ~ "1",
+    SearchReason_PossessWeapons == 1 ~ "1",
+    SearchReason_PossessEvidence == 1 ~ "1",
+    TRUE ~ "0"
+  ))
+
+cleaned_data <- cleaned_data %>%
+  rename(
+    age_group = Age_group__at_arrest_,
+    youth_at_arrest_under_18_years = Youth_at_arrest__under_18_years
+  )
+
+cleaned_data <- cleaned_data %>%
+  mutate(age_group = case_when(
+    age_group == "Aged 17 years and younger" ~ "Under 17",
+    age_group == "Aged 17 years and under" ~ "Under 17",
+    TRUE ~ age_group
+  ))
+
+cleaned_data <- cleaned_data %>%
+  mutate(booked = case_when(
+    is.na(Booked) ~ "0",
+    Booked == 1 ~ "1",
+    TRUE ~ "Not Booked"
+  ))
+
+#### Cleaned Data Conversion to Numeric ####
+
+# Convert arrest_year, search_reason, and booked to numeric
+cleaned_data <- cleaned_data |>
+  select(
+    Arrest_Year, age_group, booked, search_reason, Perceived_Race, youth_at_arrest_under_18_years) |>
+  rename(
+    arrest_year = Arrest_Year,
[…]

### 2024-09-23 Analysis of bias in arrests/scripts/00-simulate_data.R
Delete Analysis of bias in arrests directory

@@ -1,64 +1,0 @@
-#### Preamble ####
-# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
-# Author: Akshat Aneja
-# Date: 23 September 2024
-# Contact: akshat.aneja@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: none
-# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-
-#### Simulate data ####
-set.seed(1007901157)
-num_observation = 200
-# For booked arrests, which is not always 100
-num_observation_small = sample(50:100, 1, replace=TRUE)
-
-# Fictional Ethnic Identities
-sim_identities <- c("White", "East Asian", "Black", "Middle Eastern", "Latin American", "Indigenous", "South Asian", "Southeast Asian")
-
-# Fictional Age Groups
-sim_age_groups <- c("Aged 17 years and under", "Aged 18 to 24 years", "Aged 25 to 34 years", "Aged 35 to 44 years", "Aged 45 to 54 years", "Aged 55 to 64 years", "Aged 65 years and over")
-
-simulated_arrests_data <-
-  tibble(
-    "arrest_year" = sample(2020:2021, num_observation, replace = TRUE),
-    "percieved_race" = sample(sim_identities, num_observation, replace = TRUE),
-  )
-
-simulated_booked_data <-
-  tibble(
-    "arrest_year" = sample(2020:2021, num_observation_small, replace = TRUE),
-    "perceived_race" = sample(sim_identities, num_observation_small, replace = TRUE),
-  )
-
-
-simulated_age_group_data <-
-  tibble(
-    "arrest_year" = sample(2020:2021, num_observation, replace = TRUE),
-    "perceived_race" = sample(sim_identities, num_observation, replace = TRUE),
-    "age_group" = sample(sim_age_groups, num_observation, replace = TRUE),
-  )
-
-# Group and count for simulated_arrests_data
-arrests_counts <- simulated_arrests_data %>%
-  group_by(arrest_year, percieved_race) %>%
-  summarise(count = n(), .groups = "drop")
-
-# Group and count for simulated_booked_data
-booked_counts <- simulated_booked_data %>%
-  group_by(arrest_year, perceived_race) %>%
[…]

### 2024-09-23 Analysis of bias in arrests/scripts/01-download_data.R

@@ -1,33 +1,0 @@
-#### Preamble ####
-# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
-# Author: Akshat Aneja
-# Date: 23 September 2024
-# Contact: akshat.aneja@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: none
-# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
-
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-# Get package data
-
-# get package
-package <- show_package("police-race-and-identity-based-data-collection-arrests-strip-searches")
-package
-
-# get all resources for this package
-resources <- list_package_resources("police-race-and-identity-based-data-collection-arrests-strip-searches")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
-
-# load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
-
-# Save the data in CSV
-write.csv("data/raw_data/raw_data.csv", row.names = FALSE)
-
-         

### 2024-09-23 Analysis of bias in arrests/scripts/02-data_cleaning.R

@@ -1,67 +1,0 @@
-#### Preamble ####
-# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
-# Author: Akshat Aneja
-# Date: 23 September 2024
-# Contact: akshat.aneja@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: none
-# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
-
-#### Workspace setup ####
-library(tidyverse)
-
-#### Clean data ####
-data <- read.csv("data/raw_data/raw_data.csv")
-
-cleaned_data <- data |>
-  select(
-    Arrest_Year, Perceived_Race, Age_group__at_arrest_, Youth_at_arrest__under_18_years, Booked, SearchReason_CauseInjury, SearchReason_AssistEscape, SearchReason_PossessWeapons, SearchReason_PossessEvidence)
-
-cleaned_data <- cleaned_data %>%
-  mutate(search_reason = case_when(
-    SearchReason_CauseInjury == 1 ~ "1",
-    SearchReason_AssistEscape == 1 ~ "1",
-    SearchReason_PossessWeapons == 1 ~ "1",
-    SearchReason_PossessEvidence == 1 ~ "1",
-    TRUE ~ "0"
-  ))
-
-cleaned_data <- cleaned_data %>%
-  rename(
-    age_group = Age_group__at_arrest_,
-    youth_at_arrest_under_18_years = Youth_at_arrest__under_18_years
-  )
-
-cleaned_data <- cleaned_data %>%
-  mutate(age_group = case_when(
-    age_group == "Aged 17 years and younger" ~ "Under 17",
-    age_group == "Aged 17 years and under" ~ "Under 17",
-    TRUE ~ age_group
-  ))
-
-cleaned_data <- cleaned_data %>%
-  mutate(booked = case_when(
-    is.na(Booked) ~ "0",
-    Booked == 1 ~ "1",
-    TRUE ~ "Not Booked"
-  ))
-
-#### Cleaned Data Conversion to Numeric ####
-
-# Convert arrest_year, search_reason, and booked to numeric
-cleaned_data <- cleaned_data |>
-  select(
-    Arrest_Year, age_group, booked, search_reason, Perceived_Race, youth_at_arrest_under_18_years) |>
-  rename(
-    arrest_year = Arrest_Year,
[…]

### 2024-09-23 Analysis of bias in arrests/scripts/03-test_data.R

@@ -1,31 +1,0 @@
-#### Preamble ####
-# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
-# Author: Akshat Aneja
-# Date: 23 September 2024
-# Contact: akshat.aneja@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: none
-# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
-
-
-#### Workspace setup ####
-library(tidyverse)
-
-
-#### Test data ####
-arrest_data <- read.csv("data/analysis_data/analysis_data.csv")
-
-arrest_data$arrest_year == as.numeric(arrest_data$arrest_year)
-arrest_data$booked == as.numeric(arrest_data$booked)
-arrest_data$search_reason == as.numeric(arrest_data$search_reason)
-
-class(arrest_data$age_group) == "character"
-class(arrest_data$perceived_race) == "character"
-class(arrest_data$minor) == "character"
-
-booked_values <- all(arrest_data$booked >= 0 & arrest_data$booked <= 1)
-arrest_year_values <- all(arrest_data$arrest_year >= 2020 & arrest_data$arrest_year <= 2021)
-search_reason_values <- all(arrest_data$search_reason >= 0 & arrest_data$search_reason <= 1)
-
-
-

### 2024-09-23 scripts/00-simulate_data.R
Fixed file structure

@@ -1,0 +1,64 @@
+#### Preamble ####
+# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
+# Author: Akshat Aneja
+# Date: 23 September 2024
+# Contact: akshat.aneja@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(dplyr)
+
+#### Simulate data ####
+set.seed(1007901157)
+num_observation = 200
+# For booked arrests, which is not always 100
+num_observation_small = sample(50:100, 1, replace=TRUE)
+
+# Fictional Ethnic Identities
+sim_identities <- c("White", "East Asian", "Black", "Middle Eastern", "Latin American", "Indigenous", "South Asian", "Southeast Asian")
+
+# Fictional Age Groups
+sim_age_groups <- c("Aged 17 years and under", "Aged 18 to 24 years", "Aged 25 to 34 years", "Aged 35 to 44 years", "Aged 45 to 54 years", "Aged 55 to 64 years", "Aged 65 years and over")
+
+simulated_arrests_data <-
+  tibble(
+    "arrest_year" = sample(2020:2021, num_observation, replace = TRUE),
+    "percieved_race" = sample(sim_identities, num_observation, replace = TRUE),
+  )
+
+simulated_booked_data <-
+  tibble(
+    "arrest_year" = sample(2020:2021, num_observation_small, replace = TRUE),
+    "perceived_race" = sample(sim_identities, num_observation_small, replace = TRUE),
+  )
+
+
+simulated_age_group_data <-
+  tibble(
+    "arrest_year" = sample(2020:2021, num_observation, replace = TRUE),
+    "perceived_race" = sample(sim_identities, num_observation, replace = TRUE),
+    "age_group" = sample(sim_age_groups, num_observation, replace = TRUE),
+  )
+
+# Group and count for simulated_arrests_data
+arrests_counts <- simulated_arrests_data %>%
+  group_by(arrest_year, percieved_race) %>%
+  summarise(count = n(), .groups = "drop")
+
+# Group and count for simulated_booked_data
+booked_counts <- simulated_booked_data %>%
+  group_by(arrest_year, perceived_race) %>%
[…]

### 2024-09-23 scripts/01-download_data.R

@@ -1,0 +1,33 @@
+#### Preamble ####
+# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
+# Author: Akshat Aneja
+# Date: 23 September 2024
+# Contact: akshat.aneja@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
+
+
+
+#### Workspace setup ####
+library(opendatatoronto)
+library(tidyverse)
+# Get package data
+
+# get package
+package <- show_package("police-race-and-identity-based-data-collection-arrests-strip-searches")
+package
+
+# get all resources for this package
+resources <- list_package_resources("police-race-and-identity-based-data-collection-arrests-strip-searches")
+
+# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+
+# load the first datastore resource as a sample
+data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+
+# Save the data in CSV
+write.csv("data/raw_data/raw_data.csv", row.names = FALSE)
+
+         

### 2024-09-23 scripts/03-test_data.R

@@ -1,0 +1,31 @@
+#### Preamble ####
+# Purpose: Downloading and saving the data from OpenData Toronto for Police Race and Identity Based Data Collection: Arrests & Strip Searches
+# Author: Akshat Aneja
+# Date: 23 September 2024
+# Contact: akshat.aneja@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: none
+# Datasets: https://open.toronto.ca/dataset/police-race-and-identity-based-data-collection-arrests-strip-searches/
+
+
+#### Workspace setup ####
+library(tidyverse)
+
+
+#### Test data ####
+arrest_data <- read.csv("data/analysis_data/analysis_data.csv")
+
+arrest_data$arrest_year == as.numeric(arrest_data$arrest_year)
+arrest_data$booked == as.numeric(arrest_data$booked)
+arrest_data$search_reason == as.numeric(arrest_data$search_reason)
+
+class(arrest_data$age_group) == "character"
+class(arrest_data$perceived_race) == "character"
+class(arrest_data$minor) == "character"
+
+booked_values <- all(arrest_data$booked >= 0 & arrest_data$booked <= 1)
+arrest_year_values <- all(arrest_data$arrest_year >= 2020 & arrest_data$arrest_year <= 2021)
+search_reason_values <- all(arrest_data$search_reason >= 0 & arrest_data$search_reason <= 1)
+
+
+

### 2024-09-24 paper/paper.qmd
Improving abstract and checking grammar

---
title: "Analyzing Arrest Patterns and Racial Representation in Toronto"
author: "Akshat Aneja"
abstract: The city of Toronto has an extremely diverse population in terms of its ethnicity and culture. This paper examines arrest data from the Toronto police service between 2020 and 2021. It focuses on the distribution of arrests by ethnic identity, age and booking status. This is done to check whether policing practices exhibit systemic bias. By cleaning and analyzing the records, trends are explored over major ethnic groups. Despite concerns over racial bias, the data suggests that arrests are influenced by objective factors, though there still may be disparities due to socio-economic status. These findings may help in providing insights that may enhance transparency in policing practices.
date: today
date-format: long
number-sections: true
format: pdf
toc: true
thanks: "Code and data supporting this analysis is available at: https://github.com/Akshat211202/Analysis-of-bias-in-arrests/"
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(opendatatoronto)
library(dplyr)
library(tinytex)
library(ggplot2)
library(knitr)
library(kableExtra)
```

# Introduction

In recent years, debates surrounding racial bias in policing have intensified, particularly regarding its disproportionate impact on marginalized communities. This issue is especially pertinent in Toronto, a culturally and ethnically diverse city. Discussions on police reform often reflect historical connections to slavery and contemporary racial disparities, raising concerns about whether law enforcement treats marginalized communities differently, potentially leading to racial profiling.

A report from Toronto Police Services indicates that Black, East/Southeast Asian, Middle Eastern, and Latino individuals are overrepresented in use-of-force incidents relative to their population in enforcement actions. Despite efforts by Toronto Police to promote a more equitable society, these findings suggest the continued presence of racial bias in arrests.This paper analyzes arrest data from 2020 and 2021 to explore interactions between police and various ethnic groups. Disparities in arrest rates may reflect broader socio-economic factors such as poverty and unemployment, which disproportionately affect marginalized communities and contribute to higher involvement in criminal activities.

Arrest data provides a quantifiable view of law enforcement actions and their correlation with ethnicity and age. [@sec-data], outlines the data cleaning process using R, followed by an analysis of arrest distributions across age groups to identify potential trends. Age is a critical factor in this analysis, as it may reveal patterns of concern. [@sec-discussion], presents findings and discusses their implications. The study seeks to contribute to a deeper understanding of policing in Toronto, providing insights for policy reforms and community engagement efforts

# Data {#sec-data}

This data is sourced from the Toronto Open Data Portal `opendatatoronto` [@rOpenDataToronto]. The following data is used for an in-depth analysis of arrests and strip searches which involve various ethnic and identity-based groups in the city of Toronto. The information is gathered by the Toronto Police Service (TPS) [@torontoPolice] and is collected under the authority of the Police Services Act. Its purpose is to improve the relationship between the police and the community by being more transparent with the public. The data covers the period from 2020 to 2021.

The data underwent cleaning and analysis using the R programming language [@citeR]. Cleaning was performed with the `tidyverse` package [@rTidyverse], involving the removal of unnecessary columns for the analysis. Subsequently, analysis was conducted utilizing the `dplyr` package [@rDplyr], with the data grouped by ethnic and identity-based categories. Visualization was then executed using the `ggplot2` package [@rGgplot2], depicting the number of arrests and strip searches across various ethnic and identity-based groups.

The raw dataset consisted of 32,000 arrest records from 2020 and 2021, with 26 variables available for analysis. For this study, we concentrated on specific variables, including Arrest Year, Perceived Race, Age Group, Youth at Arrest, Booked, Search Reason Cause Injury, Search Reason Assist Escape, Search Reason Possess Weapons, and Search Reason Possess Evidence. To streamline the data, we combined similar columns—Search Reason Cause Injury, Search Reason Assist Escape, Search Reason Possess Weapons, and Search Reason Possess Evidence—into a single column called "Search Reason." utilizing the `any()` function [@rDplyr].

## Total number of arrests by age group for the year 2020 and 2021

It is crucial to analyze the distribution of arrests across different age groups to understand potential patterns and disparities. [@tbl-table1] presents a summary of arrest counts for various ethnic groups across all age categories in 2020 and 2021. The age groups included in the dataset are: Under 17, 18 to 24 years, 25 to 34 years, 35 to 44 years, 45 to 54 years, 55 to 64 years, and 65 years and over.

Given the complexity and extensive nature of the dataset, which involves nine age groups across eight ethnicities, summarizing the entire dataset in a comprehensible manner is challenging. To address this and ensure clarity, we have selected a subset of the data focusing specifically on individuals aged 25 to 34 years. This group was chosen as it represents a significant proportion of the data and allows for a more straightforward comparison of arrest trends across ethnic groups while avoiding information overload. By narrowing the scope, the analysis becomes more manageable and the insights more focused.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-table1
#| tbl-cap: "Arrests Booked by ethinic background for the year 2020, and 2021"

cleaned_data <- read.csv("../data/analysis_data/analysis_data.csv")

table3 <- cleaned_data %>%
  filter(!(age_group == "None") & !(perceived_race == "None")) |>
  group_by(arrest_year, perceived_race, age_group) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(perceived_race)) %>%
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

table3_small <- cleaned_data %>%
  filter(age_group == "Aged 25 to 34 years") |>
  group_by(arrest_year, perceived_race, age_group) %>%
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

kable(table3_small,
      align = "c",
      linesep = "",
      col.names = c("Ethinic and Identity Based Groups", "Age Group", "2020", "2021"),
      booktabs = TRUE)

```

## Arrests by Ethnic and Identity-Based Groups for Year 2020 and 2021

Our analysis primarily focuses on the distribution of arrests across different ethnic identities. We extracted data for the following ethnic groups: Black, White, South Asian, East Asian, Middle Eastern, Latin American, Southeast Asian, and Indigenous.

The data for 2020 and 2021 is summarized in [@tbl-table2] and [@tbl-table3]. [@tbl-table2] presents the number of individuals arrested from each ethnic group during these years, while [@tbl-table3] details how many of those arrested were subsequently booked under criminal charges.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-table2
#| tbl-cap: "Year 2020 vs Year 2021 Total number of arrests"

table1 <- cleaned_data |> 
  filter(!perceived_race == "None") |>
  group_by(arrest_year, perceived_race) |> 
  summarise(n = n(), .groups = "drop") |> 
  arrange(desc(n)) |> 
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

kable(table1,
      align = "c",
      linesep = "",
      col.names = c("Race", "2020", "2021"),
      booktabs = TRUE)
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-table3
#| tbl-cap: "Year 2020 vs Year 2021 Total number of Booked"

table2 <- cleaned_data |>
  filter(booked == 1) |>
  filter(!perceived_race == "None") |>
  group_by(arrest_year, perceived_race) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(desc(n)) |>
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

kable(table2,
      align = "c",
      linesep = "",
      col.names = c("Race", "2020", "2021"),
      booktabs = TRUE)

```

It is important to note that the OpenDataToronto [@rOpenDataToronto] provides population distribution data by ethnicity. However, due to the size and diversity of Toronto and its numerous wards, it has been challenging to cross-verify this data with crime rate statistics in a comprehensive manner. Further, a more detailed analysis could be conducted by comparing specific ethnic groups directly.

Now that the data has been cleaned, we can begin to analyze the data. The following section will provide a summary of the data, and the trends that emerge from the data. [@fig-fig1], [@fig-fig2], and [@fig-fig3], was created using `ggplot` [@rGgplot2], displays the information that can be used as compare the trends as described above.

## Total Arrests in 2020-2021

Referring to [@tbl-table2], [@tbl-table3] and [@fig-fig1] the data reveals a non-uniform distribution of arrests across ethnic groups, with the Black and White populations showing the highest numbers of arrests in both 2020 and 2021. Additionally, the data indicates a decline in arrests for these two ethnic groups from 2020 to 2021. In contrast, arrests for other ethnic groups, such as South Asian, East Asian, and Indigenous populations, increased over the same period. These opposing trends highlight the divergent patterns of arrest across different ethnic groups during this time.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "# arrests by Ethnicity in Year 2020 and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig1

table1_long <- pivot_longer(table1, cols = c("2020", "2021"), names_to = "year", values_to = "value")

# Create a plot
ggplot(table1_long, aes(x = perceived_race, y = value, fill = year)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(x = "Ethnicity",
       y = "# Arrests",
       fill = "Year") +
  theme_minimal() +
  scale_fill_discrete(labels = c("2020", "2021")) +
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Referring to [@tbl-table1] for the proportion of arrests that resulted in bookings, [@fig-fig2] was plotted with the y-axis representing the percentage of individuals booked after arrest. The data demonstrates a consistent distribution across various ethnic groups, with significantly more individuals being booked following arrest than those who were not. Notably, the percentage of individuals booked ranges between 50% and 55% across all ethnic groups, a pattern that remains consistent in 2021.

However, it is worth highlighting that the Black population had the highest number of individuals booked after arrest in both 2020 and 2021.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Booked after Arrest in the Year 2020 and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig2

table2_long <-
  cleaned_data |>
  select(
    arrest_year,
    perceived_race,
    booked
  ) |>
  group_by(arrest_year, perceived_race, booked) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(desc(perceived_race)) %>%
  filter(!(perceived_race == "None")) %>%
  pivot_wider(names_from = arrest_year, values_from = n, values_fill = 0)

table2_long <- pivot_longer(table2_long, cols = c("2020", "2021"), names_to = "year", values_to = "value")

ggplot(table2_long, aes(x = perceived_race, y = value, fill = factor(booked))) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ year, scales = "free_y") +
  labs(x = "Perceived Race",
       y = "Percentage of Arrests Booked",
       fill = "Status") +
  theme_minimal() +
  theme(legend.position = "right") + 
  scale_fill_discrete(labels = c("Not Booked", "Booked")) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")
```

## Number of Arrests by Age Group for the year 2020-2021

As depicted in [@fig-fig3], the data reveals a non-uniform distribution of arrests across age groups. The 25 to 34 and 35 to 44 age groups had the highest number of arrests in both 2020 and 2021. Additionally, the data indicates a notable decrease in arrests for these two age groups between 2020 and 2021. In contrast, other age groups experienced an increase in arrests during the same period, indicating differing trends across age demographics.

```{r}
#| warning: false
#| message: false
#| echo: false
#| eval: true
#| fig.cap: "Number of arrests by Age-Group for the year 2020, and 2021"
#| fig.width: 10
#| fig.height: 6
#| fig.align: 'center'
#| fig.pos: 'htb'
#| label: fig-fig3
# Bar plot

table3_long <- pivot_longer(table3, cols = c("2020", "2021"), names_to = "year", values_to = "value")

ggplot(table3_long, aes(x = perceived_race, y = value, fill = factor(age_group))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ year, scales = "free_y") +
  labs(x = "Perceived Race",
       y = "Number of People",
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Discussion {#sec-discussion}

While this paper offers several important insights, it is important to acknowledge its limitations. A more detailed comparison of specific ethnic groups and further research are necessary for a deeper understanding of the data. Law enforcement agencies and community advocates can use these findings to develop targeted interventions and enhance community engagement efforts. Additionally, policymakers may consider using this analysis to inform necessary policy adjustments. The need for transparency and accountability in policing remains critical, as emphasized by the Toronto Police Services Board [@rTPSB2019].

In conclusion, the analysis of Toronto's arrest data suggests that, despite concerns about racial bias, arrests appear to be influenced by objective factors rather than racial profiling. The reforms implemented by the Toronto Police Service have contributed to increased transparency and fairness within the police force. Although disparities may exist due to socio-economic factors, the data does not indicate systemic bias.
\newpage

# References

### 2024-09-24 paper/paper.qmd
Add files via upload

@@
 author: "Akshat Aneja"
-abstract: The city of Toronto has an extremely diverse population in terms of its ethnicity and culture.
-This paper examines arrest data from the Toronto police service between 2020 and 2021.
-It focuses on the distribution of arrests by ethnic identity, age and booking status.
-This is done to check whether policing practices exhibit systemic bias.
-By cleaning and analyzing the records, trends are explored over major ethnic groups.
-Despite concerns over racial bias, the data suggests that arrests are influenced by objective factors, though there still may be disparities due to socio-economic status.
-These findings may help in providing insights that may enhance transparency in policing practices.
+abstract: Toronto is known for its ethnic and cultural diversity, making it a relevant setting for examining potential biases in law enforcement.
+This paper analyzes arrest data from the Toronto Police Service for the years 2020 and 2021, focusing on the distribution of arrests by ethnic identity, age, and booking status to assess whether systemic bias exists in policing practices.
+The analysis reveals that, while arrests appear to be driven by objective factors, disparities related to socio-economic status may still exist.
+These findings provide important insights that can help improve transparency and accountability in policing, contributing to more equitable law enforcement practices.
 date: today
@@
 
-In recent years, debates surrounding racial bias in policing have intensified, particularly regarding its disproportionate impact on marginalized communities.
-This issue is especially pertinent in Toronto, a culturally and ethnically diverse city.
-Discussions on police reform often reflect historical connections to slavery and contemporary racial disparities, raising concerns about whether law enforcement treats marginalized communities differently, potentially leading to racial profiling.
+Racial bias in policing has become a central issue in public discourse, particularly regarding its disproportionate impact on marginalized communities.
+This issue is of special significance in cities like Toronto, known for its cultural and ethnic diversity.
+Amid calls for police reform, historical ties to slavery and contemporary racial disparities often dominate discussions, prompting concerns that marginalized groups may experience discriminatory treatment, including racial profiling, in interactions with law enforcement.
 
-A report from Toronto Police Services indicates that Black, East/Southeast Asian, Middle Eastern, and Latino individuals are overrepresented in use-of-force incidents relative to their population in enforcement actions.
-Despite efforts by Toronto Police to promote a more equitable society, these findings suggest the continued presence of racial bias in arrests.This paper analyzes arrest data from 2020 and 2021 to explore interactions between police and various ethnic groups.
-Disparities in arrest rates may reflect broader socio-economic factors such as poverty and unemployment, which disproportionately affect marginalized communities and contribute to higher involvement in criminal activities.
+A report from the Toronto Police Services highlights that individuals from Black, East/Southeast Asian, Middle Eastern, and Latino communities are overrepresented in use-of-force incidents compared to their presence in enforcement action populations.
+Despite efforts by the Toronto Police to promote fairness and equity, these findings suggest that racial bias may still persist in policing practices.
+However, a thorough, data-driven analysis of arrest patterns is necessary to determine the extent and nature of this bias.
+This paper addresses the gap by analyzing arrest data from 2020 and 2021 to examine how interactions between police and different ethnic groups are shaped by broader socio-economic factors such as poverty and unemployment.
+These factors, which disproportionately affect marginalized communities, may contribute to higher involvement in criminal activities, influencing arrest rates.
 
-Arrest data provides a quantifiable view of law enforcement actions and their correlation with ethnicity and age.
-[@sec-data], outlines the data cleaning process using R, followed by an analysis of arrest distributions across age groups to identify potential trends.
-Age is a critical factor in this analysis, as it may reveal patterns of concern.
-[@sec-discussion], presents findings and discusses their implications.
-The study seeks to contribute to a deeper understanding of policing in Toronto, providing insights for policy reforms and community engagement efforts
+The analysis centers on arrest data, which provides an objective and quantifiable view of law enforcement activities, focusing on the correlation between arrests, ethnicity, and age.
+The data was processed and cleaned using R, followed by an analysis of the distribution of arrests across various age groups to identify potential trends and disparities in [@sec-data].
+Age is considered a critical factor in understanding patterns of arrests, as it may reveal systemic biases in policing practices.
+The findings, discussed in [@sec-discussion], offer valuable insights into arrest patterns and their implications.
+
+This study fills a crucial gap in the understanding of racial bias in policing in Toronto by providing an empirical assessment of arrest patterns across ethnic groups.
+It seeks to contribute to ongoing discussions of police reform by offering data-driven insights that can inform policy changes and promote community engagement.
+Furthermore, this analysis highlights the importance of transparency and accountability in policing, ensuring that law enforcement actions are equitable and just for all communities.
+engagement efforts.
 

### 2024-09-26 paper/paper.qmd
Checking and improving introduction and abstract
Fixing suggestions made by peers

@@
 
-## Total number of arrests by age group for the year 2020 and 2021
+## Total number of arrests by age group for the year 2020 and 2021 {#sec-Age}
 
@@
 
-## Arrests by Ethnic and Identity-Based Groups for Year 2020 and 2021
+## Arrests by Ethnic and Identity-Based Groups for Year 2020 and 2021 {#sec-eth}
 
@@
 
-However, it is worth highlighting that the Black population had the highest number of individuals booked after arrest in both 2020 and 2021.
+However, it is worth highlighting that the Black population had the highest number of individuals who were booked after arrest in both 2020 and 2021.
 
@@
 
-## Number of Arrests by Age Group for the year 2020-2021
+## Number of Arrests by Age Group for the year 2020-2021 {#sec-group}
 
-As depicted in [@fig-fig3], the data reveals a non-uniform distribution of arrests across age groups.
-The 25 to 34 and 35 to 44 age groups had the highest number of arrests in both 2020 and 2021.
-Additionally, the data indicates a notable decrease in arrests for these two age groups between 2020 and 2021.
-In contrast, other age groups experienced an increase in arrests during the same period, indicating differing trends across age demographics.
+An analysis of arrests by age groups is essential to understanding the demographic patterns of law enforcement activity.
+[@fig-fig3] illustrates the distribution of arrests across different age groups for the years 2020 and 2021.
+The age groups included in the dataset are Under 17, 18 to 24 years, 25 to 34 years, 35 to 44 years, 45 to 54 years, 55 to 64 years, and 65 years and over.
+
+The data reveals a non-uniform distribution of arrests across age groups, with 25 to 34 and 35 to 44 age brackets accounting for the highest number of arrests in both 2020 and 2021.
+However, there is a notable decrease in the number of arrests for these age groups of 2020 to 2021 which indicates that there may be a possible shift in law enforcement activity.
+
+In contrast, the other age groups experienced an increase in arrests during the same period.
+In particular, the number of arrests for individuals aged 18 to 24 and 45 to 54 saw an upward trend from 2020 to 2021.
 
@@
 
+In [@sec-Age], the analysis highlights that arrests are not evenly distributed across age groups, with certain age ranges, particularly individuals aged 25 to 44, being more likely to be arrested during the study period.
+These findings suggest that certain demographic factors related to age may influence interactions with law enforcement.
+Further exploration is required to understand the socio-economic, behavioral, and legal factors that contribute to the higher arrest rates among individuals in their mid-20s to mid-40s.
+
+In [@sec-eth] the data suggests that there are notable disparities in arrest rates across different ethnic groups, with Black and Indigenous populations being of particular concern.
+
+In [@sec-group] the findings suggest that law enforcement practices or external factors influencing crime rates may vary significantly across age groups.
+
+However, none of these findings can definitively point to systemic bias without further context, particularly around socio-economic factors and community-specific conditions.
+
 While this paper offers several important insights, it is important to acknowledge its limitations.

### 2024-09-26 paper/paper.qmd
Adding measurement section

@@
 author: "Akshat Aneja"
-abstract: Toronto is known for its ethnic and cultural diversity, making it a relevant setting for examining potential biases in law enforcement.
-This paper analyzes arrest data from the Toronto Police Service for the years 2020 and 2021, focusing on the distribution of arrests by ethnic identity, age, and booking status to assess whether systemic bias exists in policing practices.
+abstract: This paper analyzes arrest data from the Toronto Police Service for the years 2020 and 2021, focusing on the distribution of arrests by ethnic identity, age, and booking status to assess whether systemic bias exists in policing practices.
 The analysis reveals that, while arrests appear to be driven by objective factors, disparities related to socio-economic status may still exist.
@@
 This data is sourced from the Toronto Open Data Portal `opendatatoronto` [@rOpenDataToronto].
-The following data is used for an in-depth analysis of arrests and strip searches which involve various ethnic and identity-based groups in the city of Toronto.
-The information is gathered by the Toronto Police Service (TPS) [@torontoPolice] and is collected under the authority of the Police Services Act.
-Its purpose is to improve the relationship between the police and the community by being more transparent with the public.
-The data covers the period from 2020 to 2021.
+The following data is used for an in-depth analysis of arrests and strip searches which involve various ethnic and identity-based groups in the city of Toronto covers the period from 2020 to 2021.
+The information is gathered by the Toronto Police Service (TPS) [@torontoPolice] and is collected under the authority of the Police Services Act.Instrumentation for this process involved the direct input of data by law enforcement officers during or after arrest events, ensuring the capture of relevant variables tied to each arrest.
+
+The raw dataset comprised 32,000 arrest records from 2020 and 2021, with 26 variables available for analysis which reflects real-time operational data collected through TPS's digital reporting systems, designed to improve transparency and accountability in policing practices.
+For this study, we focused on specific variables, including Arrest Year, Perceived Race, Age Group, Youth at Arrest, Booked, and the various Search Reasons (Cause Injury, Assist Escape, Possess Weapons, and Possess Evidence).
+To simplify the analysis, we consolidated these search-related columns into a single "Search Reason" column using the `any()` function [@rDplyr].
 
 The data underwent cleaning and analysis using the R programming language [@citeR].
-Cleaning was performed with the `tidyverse` package [@rTidyverse], involving the removal of unnecessary columns for the analysis.
-Subsequently, analysis was conducted utilizing the `dplyr` package [@rDplyr], with the data grouped by ethnic and identity-based categories.
-Visualization was then executed using the `ggplot2` package [@rGgplot2], depicting the number of arrests and strip searches across various ethnic and identity-based groups.
+Cleaning was performed with the `tidyverse` package [@rTidyverse].
+Subsequently, analysis was conducted utilizing the `dplyr` package [@rDplyr], visualization was done using the `ggplot2` package [@rGgplot2], depicting the number of arrests and strip searches across various ethnic and identity-based groups.
 
-The raw dataset consisted of 32,000 arrest records from 2020 and 2021, with 26 variables available for analysis.
-For this study, we concentrated on specific variables, including Arrest Year, Perceived Race, Age Group, Youth at Arrest, Booked, Search Reason Cause Injury, Search Reason Assist Escape, Search Reason Possess Weapons, and Search Reason Possess Evidence.
-To streamline the data, we combined similar columns—Search Reason Cause Injury, Search Reason Assist Escape, Search Reason Possess Weapons, and Search Reason Possess Evidence—into a single column called "Search Reason." utilizing the `any()` function [@rDplyr].
+## Measurement
+
+This study aims to estimate the extent of racial bias in arrest patterns across different ethnic groups in Toronto for the years 2020 and 2021.
+The key variables of interest include Perceived Race, Age Group, Youth at Arrest, and Booking Status.
+The primary estimand is the relationship between perceived race and arrest outcomes, specifically whether individuals from marginalized communities are disproportionately arrested or booked relative to their representation in the population.
+
+The analysis is conducted using the arrest data as the primary estimator.
+Arrest records are grouped by ethnic identity, age, and booking status to calculate the distribution of arrests across these categories.
+The study also considers broader socio-economic factors such as poverty and unemployment, which may affect arrest rates, as potential confounders.
+By quantifying these relationships through statistical analysis and data visualization, the study seeks to determine whether arrest practices reflect systemic bias or are driven by objective factors.
 
@@
   theme(legend.position = "right") + 
-  scale_fill_discrete(labels = c("Not Booked", "Booked")) +
-  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
-  scale_fill_brewer(palette = "Set1")
+  scale_fill_manual(values = c("0" = "lightblue", "1" = "darkblue"),
+                    labels = c("0" = "Not Booked", "1" = "Booked")) +
+  theme(axis.text.x = element_text(angle = 30, hjust = 1))
 ```
@@
 Although disparities may exist due to socio-economic factors, the data does not indicate systemic bias.
+
 \newpage
